---
title: "Exercise 07"
author: "Rohit Koonireddy"
date: '2022-10-30'
output: html_document
---

```{r Install Dependencies, message=FALSE, warning=FALSE, include=FALSE}
## Load essential packages in this tutorial

lib_vect <- c("raster", "rpart", "PresenceAbsence", "AUC", "randomForest", "ecospat", "biomod2", "ggplot2", "caret", "dismo", "tidyr")

get_function <- function(input) {
  if (!require(input, character.only = TRUE)) {
    install.packages(input, dependencies = c("Depends", "Suggests"))
    library(input, character.only = TRUE)
  }
}

sapply(lib_vect, get_function)
sapply(lib_vect,require,character.only=TRUE)
```

## Introduction and data set presentation
In this exercise, we will reuse the data set presented during the tutorial 5 considering the environmental variables. As you already known, the data set contains the presence and absence of bird species in Switzerland together with a few environmental variables at 1 km resolution. For this exercise, we have increased the number of bird species in the data set. The goal of this exercise is to apply Random Forest models to predict the presence absence of 18 bird species across Switzerland.  

As a reminder: 

- This data set is originally used for species distribution models (SDM). 
- SDMs use algorithms to predict the spatial distribution of species based on environmental data. 
- The environmental data is extracted from raster data cropped to the geographical region of interest.
    
You will find in this data set the following variables: 

- KoordID: a unique ID for each sampling location. It is composed of the first three figures of the x- and y-coordinates (km)
- x: Longitude coordinate of each gridcell
- y: Latitude coordinate of each gridcell

Environmental variables that have been extracted from rasters or shapefiles:

- elev_birds: elevation extracted from the gridcell, m
- prec_birds: precipitation extracted from the gridcell, mm
- temp_birds: temperature extracted from the Chelsea temperature grid, C°
- forest_edge_sum: percentage of forest edge per gridcell
- forest_sum: percentage of forest per gridcell
- grassland_sum: percentage of grassland per gridcell
- water_dist: distance to nearest body of water per gridcell
- buildings_sum: sum of buildings within the grid cell
    
From the columns 13 to the end of the data set you will find the species presence absence with the species scientific names as colnames. 
***
***
**Question 1. As a first question please explore the data set by using the code provided in the tutorial 5: **
  
  - Complete the temperature predictor by using a linear model (check the validity of the model)

```{r}
### Load the data set bird_env_extended_20.csv
# load the data set, most data you will encounter will be saved either as .txt or.csv format
bird_model_mat <- read.csv(file = "data/bird_env_extended_20.csv")

# change all "." in variable names to "_", in order to avoid confusion
colnames(bird_model_mat)
colnames(bird_model_mat) <- gsub("\\.","_", colnames(bird_model_mat))

#class data.frame
class(bird_model_mat)
#dim 427*30
dim(bird_model_mat)

#head and tail
head(bird_model_mat) 
tail(bird_model_mat) 

#check for NAs
print(paste0("NAs in Data ? : ",anyNA(bird_model_mat)))
#if there are NAs, count them per column
nas_per_column <- colSums(is.na(bird_model_mat))
columns_with_nas <- names(nas_per_column[nas_per_column > 0])
if (length(columns_with_nas) > 0) {
  cat("Columns with missing data (NAs):\n")
  print(columns_with_nas)
} else {
  cat("No missing data found in any column.\n")
}

### Complete the temperature predictor by applying a linear model 
#as we found out that elevation is best to determine this, we will do lm between temp and elev and add a column temp_pred for the predicted values
# i do this only for the values where there is NA
lm_temp <- lm(temp ~ elev, data = bird_model_mat)
predicted_temperatures <- predict(lm_temp, newdata = bird_model_mat)

bird_model_mat$temp_pred <- bird_model_mat$temp
bird_model_mat$temp_pred[is.na(bird_model_mat$temp)] <- predicted_temperatures[is.na(bird_model_mat$temp)]

head(bird_model_mat)
print(nas_per_column <- colSums(is.na(bird_model_mat)))
```

  - Plot the relation between observed and predicted temperature vs elevation 
```{r warning=FALSE}
plot_data <- data.frame(
  Observed_Temperature = bird_model_mat$temp,
  Predicted_Temperature = predicted_temperatures,
  Elevation = bird_model_mat$elev
)

ggplot(plot_data, aes(x = Elevation, y = Observed_Temperature)) +
  geom_point(aes(color = "Observed"), size = 3) +
  geom_point(aes(x = Elevation, y = Predicted_Temperature, color = "Predicted"), size = 2, shape = 3) +
  labs(x = "Elevation", y = "Temperature") +
  ggtitle("Observed vs. Predicted Temperature vs. Elevation") +
  theme_minimal() +
  scale_color_manual(values = c("Observed" = "black", "Predicted" = "blue")) +
  guides(color = guide_legend(title = "Temperature"))

```
```{r}
names(bird_model_mat)
print(nas_per_column <- colSums(is.na(bird_model_mat)))
```

  - Perform a PCA to identify the correlated variables  
```{r message=FALSE, warning=FALSE}
require(ecospat)

# Check with Pearson correlation
ecospat.cor.plot(bird_model_mat[,c("elev","prec","temp_pred","est_slope",
                                   "forest_edge_sum","forest_sum","water_dist","buildings_sum")])

#check variable importance
for(j in 13:30){
  for (i in c(4, 5, 7, 8, 9, 10, 11, 31)){
    glm.bi <- glm(bird_model_mat[,j] ~ poly(bird_model_mat[,i], 2), family = 'binomial')
    cat('Predictive Power', j, names(bird_model_mat[i]), '=', ecospat.adj.D2.glm(glm.bi)*100, '\n')
  }
  cat('\n')
}
```
According to the correlation shown in the practical and to the PCA performed here which variables will you delete?
**Solution:** 
Like in the tutorial, elevation and temperature are still highly correlated. So I would only remove elevation from the dataframe.
Final columns would be: (7 in total)
1. "prec"
2. "temp_pred"
3. "est_slope",
4. "forest_edge_sum"
5. "forest_sum"
6. "water_dist"
7. "buildings_sum"

***


**Question 2. The randomForest comes with several hyper-parameters that can be tuned. During the practical you have tested the influence of the "mtry" parameter. In this question we ask you to explore the effect of the number of tree on the model quality. First, you will have to select species with more than 10 occurrences to perform the analyses and identify the most important variables by using a glm model. Second, you will construct and fit a RandomForest model for one species: "Tetrao_tetrix". Finally, you will evaluate the model by cross validation and find the optimal number of tree that maximize the AUC value.**
  
  
 - Fit a glm model for all the species with more than 10 occurrences
```{r}
 ### Variable selection based on the first data exploration
selected_columns <- bird_model_mat[, 13:30]
column_sums <- colSums(selected_columns)
selected_columns <- selected_columns[, column_sums > 10]

selected_columns
columns_to_merge <- bird_model_mat[, c("KoordID","x","y","elev","prec","temp_pred","est_slope",
                                       "forest_edge_sum","forest_sum","grassland_sum","water_dist","buildings_sum")]

### Species selection based on the number of occurrences
more10_bird_model_mat <- cbind(columns_to_merge, selected_columns)
more10_bird_model_mat

### Fit of glm model and evaluation of the explained power of each variable for each species


```
```{r}
results <- list()

# Define the independent variables
independent_variables <- c("prec", "temp_pred", "est_slope", "forest_edge_sum", "forest_sum", "water_dist", "buildings_sum")

# Create a data frame to store the sum and average of explained powers for each independent variable
sum_and_average_of_explained_powers <- data.frame(Independent_Variable = independent_variables, 
                                                   Sum_Explained_Power = numeric(length(independent_variables)),
                                                   Average_Explained_Power = numeric(length(independent_variables)))

# Loop through each dependent variable
for (dependent_variable in names(selected_columns)) {
  #cat('Results for', dependent_variable, ':\n')

  # Create a data frame to store results for the current dependent variable
  dependent_results <- data.frame(Independent_Variable = independent_variables, Explained_Power = numeric(length(independent_variables)))

  # Loop through each independent variable
  for (i in 1:length(independent_variables)) {
    ind_var <- independent_variables[i]
    #cat('  Independent Variable:', ind_var, '\n')

    # Create a formula for the logistic regression model
    formula <- as.formula(paste(dependent_variable, "~", ind_var))

    # Fit a glm model for the current dependent and independent variable
    glm_model <- glm(formula, family = 'binomial', data = bird_model_mat)

    # Calculate explained power for the independent variable
    explained_power <- ecospat.adj.D2.glm(glm_model) * 100

    # Store the results for the independent variable in the data frame
    dependent_results[i, "Explained_Power"] <- explained_power
  }

  # Store the results for the current dependent variable in the results list
  results[[dependent_variable]] <- dependent_results

  # Update the sum of explained powers and compute the average
  sum_and_average_of_explained_powers$Sum_Explained_Power <- sum_and_average_of_explained_powers$Sum_Explained_Power + dependent_results$Explained_Power
}

# Calculate the average of explained powers for each independent variable
sum_and_average_of_explained_powers$Average_Explained_Power <- sum_and_average_of_explained_powers$Sum_Explained_Power / length(names(selected_columns))

# View the results as a list of data frames for each dependent variable
#results

# View the sum and average of explained powers for each independent variable
sum_and_average_of_explained_powers

```
 

 - Check for variable importance and select the variable with more that 2% of explained power
water_dist has an average explained power than 2% as seen above. Sum of explained power is not relevant. 
```{r , warning=FALSE}
 ### Delete the variables with less than 2% of explained power that you do not want to include in the model
 
more10_bird_model_mat <- more10_bird_model_mat[, !(names(more10_bird_model_mat) == "water_dist")]
names(more10_bird_model_mat)
```
- Construct the RandomForest model for one species: "Tetrao_tetrix"
```{r,message=FALSE}
library(randomForest)

independent_variables <- c("prec", "temp_pred", "est_slope", "forest_edge_sum", "forest_sum", "buildings_sum")
species_to_model <- "Tetrao_tetrix"
more10_bird_model_mat[,species_to_model] <- as.factor(more10_bird_model_mat[,species_to_model])

formula <- as.formula(paste(species_to_model, "~", paste(independent_variables, collapse = " + ")))
rf_model <- randomForest(formula, data = more10_bird_model_mat, importance = TRUE, type = "classification")

print(rf_model)
```

-  Find the optimal number of trees. To do so use the provided cross validation function.
```{r,message=FALSE }
# cross-validation in random forest
```


```{r,message=FALSE }
library(dismo)
cv.model.rf <- function(model, K, data, spp,independent_variables){
  data <- data[,c(independent_variables,spp)]
  ks <- kfold(data, k = K, by = data[,spp])
  cvpreds <- data.frame(row = row.names(data),observed=data[,spp], cvpred = numeric(length = nrow(data)))
  AUC_sum <- data.frame(matrix(NA,ncol = 2,nrow = K))
  names(AUC_sum) <- c("train_AUC","test_AUC")
  
  
   for(i in 1:K){
    train <- data[ks!=i,]
    test <- data[ks==i,]
    modtmp <- update(model, data = train)
    cvpreds[which(ks==i),3] <- predict(modtmp, newdata = test, type = 'prob')[,2] 
    train_pred <- predict(modtmp, newdata = train, type = 'prob')[,2]
    test_pred <- predict(modtmp, newdata = test, type = 'prob')[,2]
    
    AUC_sum[i,1] <- AUC::auc(AUC::roc(train_pred, as.factor(train[,spp])))
    AUC_sum[i,2] <- AUC::auc(AUC::roc(test_pred, as.factor(test[,spp])))
  }
cvpreds
return(c(Training=mean(AUC_sum$train_AUC),Test=mean(AUC_sum$test_AUC)))
}
```

-  To optimize the calculation please use the following sequence for the number of tree.
```{r include=FALSE}
Nb_tree <- c(seq(1,100,2),seq(200,1000,25))
```

-  Plot the AUC values for both the training and the test data sets against the number of trees on the same plot.
```{r warning=FALSE}
### make the plot
train_AUC_values <- numeric(length(Nb_tree))
test_AUC_values <- numeric(length(Nb_tree))

for (i in seq_along(Nb_tree)) {
  n_trees <- Nb_tree[i]
  model_rf <- randomForest(formula, data = more10_bird_model_mat, ntree = n_trees, importance = TRUE, type = "classification")
  
  auc_values <- cv.model.rf(model_rf, K = 5, data = more10_bird_model_mat, spp = species_to_model, independent_variables)
  
  train_AUC_values[i] <- auc_values[1]
  test_AUC_values[i] <- auc_values[2]
}
```
```{r}
#plot

# Get the tree with the highest AUC
best_tree_idx <- which.max(test_AUC_values)
best_tree <- getTree(rf_model, k = best_tree_idx, labelVar = TRUE)
best_nb_tree <- Nb_tree[best_tree_idx]
print(paste("Best trees: ",best_nb_tree))

#plot the graphs
plot(Nb_tree, train_AUC_values, type = "l", col = "blue", xlab = "Number of Trees", ylab = "AUC", main = "AUC vs. Number of Trees")
lines(Nb_tree, test_AUC_values, type = "l", col = "red")
legend("topright", legend = c("Training AUC", "Test AUC"), col = c("blue", "red"), lty = 1)
```

***
**Question 3. As you have defined a minimum number of tree in the previous question, you can now construct a function that help to fit and evaluate RandomForest model for all species.**  This function needs to take as input arguments:

- the bird data set (that encompass the coordinates of each occurrence, the ID of each cells, the environmental variables and the species names).
- the species names
- the selected variables

The function needs to return the RF model, the model prediction, the model evaluation (kappa, TSS, the sensitivity, the specificity and the AUC and the cross validation test). Test your function with the species: "Tetrao_tetrix".
```{r warning=FALSE, message=FALSE}
# Load required libraries
library(randomForest)
library(caret)
library(ROCR)
```

```{r}
names(more10_bird_model_mat)
```
```{r}
library(dismo)

cv.model.rf.NEW <- function(local_model, K, data, spp, independent_variables, num_trees = 525, local_formula) {
  data <- data[, c(independent_variables, spp)]
  ks <- kfold(data, k = K, by = data[, spp])
  print(paste("spp:", spp))
  cvpreds <- data.frame(
    row = row.names(data),
    observed = data[, spp],
    cvpred = numeric(length = nrow(data))
  )
  
  AUC_sum <- data.frame(matrix(NA, ncol = 4, nrow = K))
  names(AUC_sum) <- c("train_AUC", "test_AUC", "sensitivity", "specificity")
  
  for (i in 1:K) {
    train <- data[ks != i, ]
    test <- data[ks == i, ]
    modtmp_local <- update(local_model, data = train)
    cvpreds[which(ks == i), 3] <- predict(modtmp_local, newdata = test, type = 'prob')[, 2]
    train_pred <- predict(modtmp_local, newdata = train, type = 'prob')[, 2]
    test_pred <- predict(modtmp_local, newdata = test, type = 'prob')[, 2]
    
    AUC_sum[i, 1] <- AUC::auc(AUC::roc(as.numeric(train_pred), as.factor(train[, spp])))
    AUC_sum[i, 2] <- AUC::auc(AUC::roc(as.numeric(test_pred), as.factor(test[, spp])))
    
    # Calculate sensitivity and specificity
    confusion_matrix <- confusionMatrix(table(Predicted = ifelse(test_pred > 0.5, "1", "0"), Actual = as.factor(test[, spp])))
    sensitivity <- confusion_matrix$byClass["Sensitivity"]
    specificity <- confusion_matrix$byClass["Specificity"]
    
    AUC_sum[i, 3] <- sensitivity
    AUC_sum[i, 4] <- specificity
  }

  # Calculate mean AUC values, sensitivity, specificity
  train_mean_auc <- mean(AUC_sum$train_AUC)
  test_mean_auc <- mean(AUC_sum$test_AUC)
  mean_sensitivity <- mean(AUC_sum$sensitivity)
  mean_specificity <- mean(AUC_sum$specificity)

  return(list(
    Training = train_mean_auc,
    Test = test_mean_auc,
    Sensitivity = mean_sensitivity,
    Specificity = mean_specificity,
    TSS = mean_sensitivity+mean_specificity-1,
    ConfusionMatrix = confusion_matrix
  ))
}
```

```{r}
# Define the function to fit and evaluate RandomForest models
get_rF <- function(given_bird_data, species_names, selected_variables, number_trees) {
  results <- list()
  for (local_species in species_names) {
    print(paste("Fitting RandomForest model for species:", local_species))
    bird_data <-  given_bird_data[,c(local_species,selected_variables)]
    bird_data[,local_species] <- as.factor(given_bird_data[,local_species])
    local_formula <- as.formula(paste(local_species, "~", paste(selected_variables, collapse = " + ")))
    rf_model_local <- randomForest(local_formula, data = bird_data, importance = TRUE, type = "classification",num_trees= number_trees)
    cv_results <- cv.model.rf.NEW(local_model = rf_model_local, K = 5, data = bird_data, 
                                  spp = local_species, independent_variables = selected_variables, local_formula = local_formula)
    result <- list(
      CrossValidation <- cv_results
    )

    results[[local_species]] <- result
  }

  return(results)
}
```


```{r}
# Test the function with the species "Tetrao_tetrix" as an example
needed_species <- c("Tetrao_tetrix")
independent_variables <- c("prec", "temp_pred", "est_slope", "forest_edge_sum", "forest_sum", "buildings_sum")
data_birds <- more10_bird_model_mat  # Load your bird data here

results <- get_rF(given_bird_data = data_birds, species_names = needed_species, selected_variables = independent_variables)

# Print the results for "Tetrao_tetrix"
print(results)
```
```{r}
needed_species <- names(more10_bird_model_mat)[12:20]
independent_variables <- c("prec", "temp_pred", "est_slope", "forest_edge_sum", "forest_sum", "buildings_sum")
data_birds <- more10_bird_model_mat

### Apply the function to all the species
results <- get_rF(given_bird_data = data_birds, species_names = needed_species, selected_variables = independent_variables)
```

***

**Question 4. Apply the function to all the species present into the data set and select only the model that present a good fit to the observed data (TSS > 0.6). We also want to project the model to whole Switzerland to get a map of the species' distribution. For this you need to load a raster stack that includes the environmental variables for whole Switzerland.**
```{r warning=FALSE, message=FALSE,eval=FALSE}
### Load the raster stack and rename the layers to match the model matrix
bird_extent_stack <- readRDS('data/bird_extent_stack.rds')
bird_extent_stack
```
```{r}
library(raster)

# Read the raster stack from your RDS file
bird_extent_stack <- readRDS('data/bird_extent_stack.rds')
layer_has_nas <- is.na(bird_extent_stack)
bird_extent_df <- as.data.frame(bird_extent_stack, xy = TRUE, na.rm = TRUE)
colnames(bird_extent_df) <- gsub("\\.","_", colnames(bird_extent_df))
colnames(bird_extent_df) <- c("x","y","elev","est_slope","temp_pred","prec","forest_edge_sum","forest_sum", "grassland_sum","water_dist","buildings_sum")

selected_best_species <- "Parus_caeruleus"
independent_variables <- c("prec", "temp_pred", "est_slope", "forest_edge_sum", "forest_sum", "buildings_sum")

# Subset the environmental data for the selected species
condensed_bird_extent_df <- bird_extent_df[, independent_variables]

# Ensure the response variable is a factor
bird_model_mat[, selected_best_species] <- as.factor(bird_model_mat[, selected_best_species])

best_formula <- as.formula(paste(selected_best_species, "~", paste(independent_variables, collapse = " + ")))

best_model <- randomForest(best_formula, data = more10_bird_model_mat, importance = TRUE, type = "classification", ntree = 525)

# If your dataset includes NA values, you might want to use the na.action argument to handle them, for example:
best_model_predictions <- predict(best_model, condensed_bird_extent_df, type = "response")
predicted_classes <- ifelse(best_model_predictions >= 0.5, 1, 0)

value_counts <- table(predicted_classes)
print(value_counts)

```
***

**Question 5. Finally, compare the results obtained by fitting the RF model to the observed data and discuss the results. What could be the next step?**
- Plot and compare the observed map and the predicted 
I do not do this solely because I could not compare with the actual data in given raster if you can find the field for actual predicitons, it is only few lines of code.
```{r eval=FALSE, fig.dim=, include=FALSE}
### Plot the observed vs predicted values for for the following species

### Transform the data into Raster
# Convert observed and predicted data to Raster objects

### Plot the observed vs predicted values for the first 3 species
#levelplot(observed_raster, main = "Observed Map")  # Adjust as needed
#levelplot(predicted_raster, main = "Predicted Map")  # Adjust as needed
```

**Solution:** 

